Efficient implementation of kernel filtering

### 2007-06-25 article/Article.tex
Corrections to article
darcs-hash:20070625044023-fafb9-4cc85177df113c5637fc39f7f1a9b7fcf1cdce9b.gz

\begin{abstract}
\noindent
% The abstract should be a paragraph or two long, and describe the
% scope of the document.
Kernel based filtering is one of the fundamental tools of image
analysis and processing. A number of approaches have been developed
that allow efficient implementation of such filters even when the
kernel size is large. This article reviews some of these methods and
provides introduces their ITK implementations.
\end{abstract}

@@

\section{Introduction}
A kernel based filtering process replaces the pixel at the kernel
origin with the result of applying a function to all pixels defined by
the kernel. Many useful filters, including edge detection and gradient
filters, smoothing filters and rank and morphology filters fall into
this category. Direct implementations of such filters typically
involve visiting all pixels defined by the kernel in order to evaluate
the filter function. Such an approach is usually simple to implement
-- in ITK it is made simple by the neighborhood iterators -- but leads
to an algorithm complexity proportional to the number of pixels in the
kernel (or $O(n^d)$, where $n$ is the kernel size and $d$ is the
dimenionality). Such complexity tends to restrict application of such
filters to small kernels.

A number of classical methods exist for reducing this complexity to
more managable, in some cases kernel size independent, levels. ITK
already exploits such techniques for gaussian convolution
operations. This paper describes the classical techniques for
optimized mathematical morphology filters, rank filters and certain
convolution filters. These filters make the use of large kernels,
which can be very useful in many applications, practical on
conventional computing hardware.

\section{Separability and recursive implementations}
The two approaches most typically used to reduce complexity of kernel
based filters are separability and recursive computation, both of
which are used in the ITK implementation of gaussian convolution
filters. A separable filter implements a multidimensional kernel by
cascading several one dimensional kernels, therefore reducing
complexity from $O(n^d)$ to $O(nd)$. The second approach exploits
redundancy that might be present in the computations of kernel
functions at neighboring locations, leading, in some cases, to a complexity
independent of $n$.

\section{Mathematical morphology operations}
Two optimized forms of the mathematical morphology operations of
erosion, dilation, opening and closing are presented here. The kernel
is usually referred to as the structring element in mathematical
morphology. The methods described here are applied to ``flat''
structuring elements, that is structuring elements without
weights. The first method can be applied to arbitary structuring
elements while the second can be applied to line structuring elements.

\subsection{Arbitary structuring elements}
\label{sect:MMmovingHist}
The method described in \cite{Vandroogenbroeck96.3} relies on the
simple concept of an updatable histogram, and is often described as a
``moving histogram'' approach. A histogram is computed for a kernel
located at the first voxel. The histogram at the neighboring voxel can
then be computed by including newly included voxels and removing newly
excluded voxels. The list of included and excluded voxels
corresponding to movement in any direction can be computed when the
structuring element is created, and the direction with the smallest
number of changes should be selected as the direction for sweeping the
kernel across the image. The erosion or dilation at each location is
computed by selecting the minimum or maximum from the histogram. This
approach is very efficient when 8 or 16 bit pixels are used because
the histogram can be represented as an array, with place holders
tracking the current maximum or minimum increasing performance. More
sophisticated histogram representations are necessary for larger pixel
types. Our implementation uses c++ maps. This is an extension to the
original paper.

@@

\subsection{Decomposition of structuring elements}
\label{sect:MMdecomp}
Morphological erosions and dilations are separable -- successive
dilations by orthogonal lines is equivalent to dilation by a rectangle
with sides equal to the line lengths. This means that any
hyper-rectangular structuring element can be constructed using lines
parallel to the image axes.

Approximations of more complex shapes, particulary circles and
ellipses can constructed using a number of lines at evenly spaced
angles \cite{Adams93}. It is difficult to create a structuring element
with a precisely defined radius using this method because line
structuring elements from which the circle structuring element is
composed must have odd length and there are practical limits due to
the realities of underlying digital grid representation of images. In
addition it is possible that the structuring elements may not be truly
translation invariant due to the representation of line used
(e.g. Bresenham) in the decomposition. However, precisely defined
radii are rarely critical when a large structuring element is called
for. An example of a structuring element created using line
structuring elements is shown in Figure \ref{fig:circledecomposition}. Composition of regular
shapes, such as hexagons and octagons is more accurate.

@@

It is also possible, in theory, to construct 3D structuring elements
in similar ways. The construction of a hyper-rectangle is trivial,
however construction of spheres is more problematic. The code
discussed later provides preliminary implementations based on some
platonic solids and various spherical approximateions, but further
testing and development is needed.

@@

A method published recently that offers improved
performance \cite{Vandroogenbroeck2005Morphological} and the ability
to perform a line opening directly (rather than using an
erosion/dilation cascade). This method has been implemented in the
filters discussed later in the article, but performance issues are not
yet clear. The method employs histograms and therefore needs more
complex data structures when applied to higher precision data,
potentially any speed advantage.

\section{Rank filters}
\label{sect:rank}
Efficient implemetations of median and rank filters can be carried out
using exactly the same approach as discussed for morphological
operations with arbitary structuring elements. The method was
originally proposed in \cite{Huang79}. The implementation discussed
later supports arbitary kernel shapes and pixel types as well as any
choice of rank.

Rank filters are not separable. However the performance benefits
offered by separability make it worth pretending they are. If median
filtering is being used to provide robust noise filtering or
background estimation then a separable approximation is worth
testing. This concept was originally proposed in \cite{Narendra81}. In
fact it is also posible to apply the decomposition discussed in
Section \ref{sect:MMdecomp} to median filters to achieve a closer
approximation to circular kernels.

@@

A more efficient strategy was proposed in \cite{Crow84} for rectangular
kernels. This approach first computes an accumulator image in which
each $i,j$ is repaced by .. This accumulator image may be computed
recursively using local neighborhood values. The mean of rectangles of
any size may then be computed using accumulated values at the
rectangle corners. Figures \ref{fig:accum} amd \ref{fig:accumMean}
illustrate the 2D case and the update formulas are provided in
Equations \ref{eq:accum} and \ref{eq:accumMean}.

@@

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.5]{accumMean2d}
\caption{Computation of mean using the accumulated image - lue at $(i,j)$ computed using Equation \ref{eq:accumMean}.\label{fig:accumMean}}
\end{figure}

@@

\subsection{Rank and mean filters -- fastRankMean}
The {\em fastRankMean} packge using the sliding window method to
implement rank and mean filters for arbitary kernel shapes. It also
uses line versions of these kernels to provide a ``separable rank''
filter -- an fast approximation to a real rank filter as discussed in
Section \ref{sect:rank}. There are also a number of filters that
accept a mask input and return the rank or mean value of the
intersection of kernel and mask. These are experimental and haven't
been thoroughly tested.

\subsection{Rectangular convolution -- boxConvolution}
The {\em boxConvolution} uses the accumulation method described in
Section \ref{sect:meanVar} to implement rectangular mean and variance
computations. The filters implement different boundary conditions to
{\em itkMeanImageFilter}, with the mean of the kernel inside the image
being computed.

@@

>>>>

\begin{abstract}
\noindent
% The abstract should be a paragraph or two long, and describe the
% scope of the document.
Kernel based filtering is one of the fundamental tools of image
analysis and processing. A number of approaches have been developed
over the years that allow efficient implementation of such filters
even when the kernel size is large. This article reviews some of these
methods and introduces their ITK implementations.
\end{abstract}

@@

\section{Introduction}
A kernel based filtering process replaces the pixel at the kernel
origin with the result of applying a function to all pixels defined by
the kernel. Many useful filters, including edge detection and gradient
filters, smoothing filters and rank and morphology filters fall into
this category. Direct implementations of such filters typically
involve visiting all pixels defined by the kernel in order to evaluate
the filter function. Such an approach is usually easy to implement
-- in ITK it is made simple by the neighborhood iterators -- but leads
to an algorithm complexity proportional to the number of pixels in the
kernel (or $O(n^d)$, where $n$ is the kernel size and $d$ is the
dimensionality). Such complexity tends to restrict application of such
filters to small kernels.

A number of classical methods exist for reducing this complexity to
more manageable, in some cases kernel size independent, levels. ITK
already exploits such techniques for Gaussian convolution
operations. This paper describes the classical techniques for
optimized mathematical morphology filters, rank filters and certain
convolution filters. These filters make the use of large kernels,
which can be very useful in many applications, practical on
conventional computing hardware.

\section{Separability and recursive implementations}
The two approaches most typically used to reduce complexity of kernel
based filters are separability and recursive computation, both of
which are used in the ITK implementation of Gaussian convolution
filters. A separable filter implements a multidimensional kernel by
cascading several one dimensional kernels, therefore reducing
complexity from $O(n^d)$ to $O(nd)$. The second approach exploits
redundancy that might be present in the computations of kernel
functions at neighboring locations, leading, in some cases, to a complexity
independent of $n$.

\section{Mathematical morphology operations}
Two optimized forms of the mathematical morphology operations of
erosion, dilation, opening and closing are presented here. The kernel
is usually referred to as the structuring element in mathematical
morphology. The methods described here are applied to ``flat''
structuring elements, that is structuring elements without
weights. The first method can be applied to arbitrary structuring
elements while the second can be applied to line structuring elements.

\subsection{Arbitrary structuring elements}
\label{sect:MMmovingHist}
The method described in \cite{Vandroogenbroeck96.3} relies on the
simple concept of an updatable histogram, and is often described as a
``moving histogram'' approach. A histogram is computed for a kernel
located at the first voxel. The histogram at the neighboring voxel can
then be computed by including newly included voxels and removing newly
excluded voxels. The list of included and excluded voxels
corresponding to movement in any direction can be computed when the
structuring element is created, and the direction with the smallest
number of changes should be selected as the direction for sweeping the
kernel across the image. The erosion or dilation at each location is
computed by selecting the minimum or maximum from the histogram. This
approach is very efficient when 8 or 16 bit pixels are used because
the histogram can be represented as an array, with place holders
tracking the current maximum or minimum increasing performance. More
sophisticated histogram representations are necessary for larger pixel
types. Our implementation uses c++ maps. This is an extension to the
original paper.

@@

\subsection{Decomposition of structuring elements}
\label{sect:MMdecomp}
Morphological erosions and dilations are separable -- successive
dilation by orthogonal lines is equivalent to dilation by a rectangle
with sides equal to the line lengths. This means that any
hyper-rectangular structuring element can be constructed using lines
parallel to the image axes.

Approximations of more complex shapes, particularly circles and
ellipses can constructed using a number of lines at evenly spaced
angles \cite{Adams93}. It is difficult to create a structuring element
with a precisely defined radius using this method because line
structuring elements from which the circle structuring element is
composed must have odd length and there are practical limits due to
the realities of underlying digital grid representation of images. In
addition it is possible that the structuring elements may not be truly
translation invariant due to the representation of line
(e.g. Bresenham) used in the decomposition. However, precisely defined
radii are rarely critical when a large structuring element is called
for. An example of a structuring element created using line
structuring elements is shown in Figure
\ref{fig:circledecomposition}. Composition of regular shapes, such as
hexagons and octagons is more accurate.

@@

It is also possible, in theory, to construct 3D structuring elements
in similar ways. The construction of a hyper-rectangle is trivial,
however construction of spheres is more problematic. The code
discussed later provides preliminary implementations based on some
platonic solids and various spherical approximations, but further
testing and development is needed.

@@

A method published recently that offers improved performance
\cite{Vandroogenbroeck2005Morphological} and the ability to perform a
line opening directly (rather than using an erosion/dilation
cascade). This method, known as {\em anchor morphology} has been
implemented in the filters discussed later in the article, but
performance issues are not yet clear. The method employs histograms
and therefore needs more complex data structures when applied to
higher precision data, potentially reducing any speed advantage.

\section{Rank filters}
\label{sect:rank}
Efficient implementations of median and rank filters can be carried out
using exactly the same approach as discussed for morphological
operations with arbitrary structuring elements. The method was
originally proposed in \cite{Huang79}. The implementation discussed
later supports arbitrary kernel shapes and pixel types as well as any
choice of rank.

Rank filters are not separable. However the performance benefits
offered by separability make it worth pretending they are. If median
filtering is being used to provide robust noise filtering or
background estimation then a separable approximation is worth
testing. This concept was originally proposed in \cite{Narendra81}. In
fact it is also possible to apply the decomposition discussed in
Section \ref{sect:MMdecomp} to median filters to achieve a closer
approximation to circular kernels.

@@

A more efficient strategy was proposed in \cite{Crow84} for
rectangular kernels. This approach first computes an accumulator image
in which each $i,j$ is replaced by the sum of voxels ``left'' and
``above'' it. This accumulator image may be computed recursively using
local neighborhood values. The mean of rectangles of any size may then
be computed using accumulated values at the rectangle corners. Figures
\ref{fig:accum} and \ref{fig:accumMean} illustrate the 2D case and the
update formulas are provided in Equations \ref{eq:accum} and
\ref{eq:accumMean}.

@@

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.5]{accumMean2d}
\caption{Computation of mean using the accumulated image - value at $(i,j)$ computed using Equation \ref{eq:accumMean}.\label{fig:accumMean}}
\end{figure}

@@

The availability of line morphology operations means that new
operations, such as ``union of openings'' can be implemented easily.

\subsection{Rank and mean filters -- fastRankMean}
The {\em fastRankMean} packge uses the sliding window method to
implement rank and mean filters for arbitary kernel shapes. It also
uses line versions of these kernels to provide a ``separable rank''
filter -- an fast approximation to a real rank filter as discussed in
Section \ref{sect:rank}. There are also a number of filters that
accept a mask input and return the rank or mean value of the
intersection of kernel and mask. These are experimental and haven't
been thoroughly tested.

\subsection{Rectangular convolution -- boxConvolution}
The {\em boxConvolution} uses the accumulation method described in
Section \ref{sect:meanVar} to implement rectangular mean and variance
computations. The filters implement different boundary conditions to
{\em itkMeanImageFilter}, with the mean of the kernel inside the image
being computed. The filters in this package are multithreaded and deal
with arbitary dimensions.

@@

\section{Comments and Conclusions}
